---
title: "DATA 624 HW1"
author: "Josh Iden"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    ##usebookdown: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
# load packages 
library(tidyverse)
library(fpp2)
library(gridExtra)
```

# HA 2.1

#### Use the help function to explore what the series `gold`, `woolyrnq` and `gas` represent.

```{r}
help('gold')
```

Time series data containing daily morning gold prices in US dollars, 1 January 1985 - 31 March 1989.

```{r}
help('woolyrnq')
```

Time series data containing quarterly production of woollen yarn in Australia: tonnes. Mar 1965 - Sep 1994. Source: Time Series Data Library. <https://pkg.yangzhuoranyang.com/tsdl/>

```{r}
help('gas')
```

Time series data containing Australian monthly gas production: 1956-1995. Source: Australian Bureau of Statistics.

#### a) Use autoplot() to plot some of the series in these data sets.

```{r, cache=TRUE, fig.height=15}
p1 = autoplot(gold) + ggtitle('Price of Gold: 1985-1989') + ylab('price') + xlab('')
p2 = autoplot(woolyrnq) + ggtitle('Quarterly Production of Woollen Yarn in Australia: 1965 - 1994') + ylab('tonnes') + xlab('')
p3 = autoplot(gas) + ggtitle('Australian monthly gas production: 1956-1995') + ylab('gas') + xlab('')

grid.arrange(p1, p2, p3, nrow=3)
```

#### b) What is the frequency of each series? Hint: apply the `frequency()` function.

```{r}
sprintf('gold frequency = %d', frequency(gold))
sprintf('wool frequency = %d', frequency(woolyrnq))
sprintf('gas frequency = %d', frequency(gas))
```

Gold data is yearly, Wool data is quarterly, and Gas data is monthly.

#### c) Use `which.max()` to spot the outlier in the `gold` series. Which observation was it?

```{r}
cat(sprintf('day: %d\nprice: $%1.2f', which.max(gold), gold[which.max(gold)]))
```

The outlier is at observation day 770. The price of gold on this day was \$593.70.

# HA 2.3

#### Download some monthly Australian retail data from [the book website](https://otexts.com/fpp2/extrafiles/retail.xlsx). These represent retail sales in various categories for different Australian states, and are stored in a MS-Excel file.

#### You can read the data into R with the following script:

```{r, cache=TRUE}
retaildata = readxl::read_excel('retail.xlsx', skip = 1)
```

#### The second argument (skip = 1) is required because the Excel sheet has two header rows.

#### Select one of the time series as follows (but replace the column name with your own chosen column):

```{r, cache=TRUE}
myts = ts(retaildata[,'A3349396W'],
          frequency = 12, start = c(1982, 4))
```

#### Explore your chosen retail time series using the following functions:

`autoplot()`, `ggseasonplot()`, `ggsubseriesplot()`, `gglagplot()`, `ggAcf()`.

#### Can you spot any seasonality, cyclicity and trend? What do you learn about the series?

```{r, cache=TRUE, fig.height=30}
auto = autoplot(myts) + ggtitle('lineplot')
season = ggseasonplot(myts) + ggtitle('season plot')
subseries = ggsubseriesplot(myts) + ggtitle('subseries plot')
lag = gglagplot(myts) + ggtitle('lagplot') 
acf = ggAcf(myts) + ggtitle('ACF plot')

grid.arrange(auto, season, subseries, lag, acf, nrow=5)
```

We can see from the **line plot** that there is a clear, increasing trend, and there is some seasonality. The **lagged scatterplots** show that the relationships are strongly positive across all lags, but dramatically positive at lag 12. This indicates a high annual correlation. We can see this in the **ACF** plot as well. We also notice that the correlation, while remaining positive, becomes less strongly positive as time goes by, and then peaks again every 12 months. The **subseries** plot shows us that December has the highest average monthly sales and lowest in February, which we can see from the **seasonal** plot is a trend that is consistent across years.

# HA 6.2

#### The `plastics` data set consists of monthly sales (in thousands) of product A for a plastics manufacturer for five years.

#### a) Plot the time series of sales of product A. Can you identify seasonal fluctuations and/or a trend cycle?

```{r, cache=TRUE}
autoplot(plastics) + ggtitle('Monthly Sales of Product A') + ylab('Sales (in Thousands)') + xlab('Year')
```

We can see that sales are highest midway through the year and lowest at the beginning and end of the year. A seasonal plot can give us another view of this,

```{r, cache=TRUE}
ggseasonplot(plastics) + ggtitle('Monthly Sales of Product A') + ylab('Sales (in Thousands)') + xlab('Year')
```

Interestingly, we can see that in years 1-4, sales peaked later in the year, while sales peaked earlier in year 5. We can see there is an increasing trend from years 1-5, but in year 5 sales dropped below the levels of years 3-4.

#### b) Use a classical multiplicative decomposition to calculate the trend-cycle and seasonal indices.

```{r, cache=TRUE}
plastics |>
  decompose(type = 'multiplicative') |>
  autoplot() + xlab('Year') +
  ggtitle('Classical multiplicative decomposition of Product A index')
```

The remainder values below 1 indicate that there is some "leakage" of the trend-cycle component into the remainder component - the trend-cycle estimate has over-smoothed the drop in data.

#### c) Do the results support the graphical interpretation from part a?

Yes, insofar as the trend over time can be observed while identifying the dip at the end of year 5. The seasonality of sales dropping at the beginning/end of each year and peaking during the middle of the year is also supported in part b.

#### d) Compute and plot the seasonally adjusted data

```{r, cache=TRUE}
fit <- plastics |>
  decompose(type = 'multiplicative')

autoplot(plastics, series='Data') +
  autolayer(seasadj(fit), series='Seasonally Adjusted') +
  xlab('Year') + ylab('Sales (in Thousands)') +
  ggtitle('Monthly Sales of Product A') +
  scale_color_manual(values=c('grey','red'),
                     breaks=c('Data','Seasonally Adjusted'))
```

#### e) Change one observation to be an outlier (e.g. add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?

```{r, cache=TRUE}
# generate random number
set.seed(123)
index <- sample(1:length(plastics), 1)
# generate copy of data 
plastics.copy <- plastics
# add 500 to the random index 
plastics.copy[index] <- plastics.copy[index] + 500

# recompute seasonally adjusted data 
fit.copy <- plastics.copy |>
  decompose(type = 'multiplicative')

autoplot(plastics.copy, series='Data') +
  autolayer(seasadj(fit.copy), series='Seasonally Adjusted') +
  xlab('Year') + ylab('Sales (in Thousands)') +
  ggtitle('Monthly Sales of Product A (with Outlier)') +
  scale_color_manual(values=c('grey','red'),
                     breaks=c('Data','Seasonally Adjusted'))
```

The outlier affects the Seasonally Adjusted data as consistent with its effect to the main data.

#### e) Does it make any difference if the outlier is near the end rather than in the middle of the time series?

```{r, cache=TRUE}
# set later index
set.seed(123)
index.end <- sample(55:60, 1)

# generate copy of data 
plastics.copy.v2 <- plastics
# add 500 to the random index 
plastics.copy.v2[index.end] <- plastics.copy.v2[index.end] + 500

# recompute seasonally adjusted data 
fit.copy.v2 <- plastics.copy.v2 |>
  decompose(type = 'multiplicative')

autoplot(plastics.copy.v2, series='Data') +
  autolayer(seasadj(fit.copy.v2), series='Seasonally Adjusted') +
  xlab('Year') + ylab('Sales (in Thousands)') +
  ggtitle('Monthly Sales of Product A (with Outlier)') +
  scale_color_manual(values=c('grey','red'),
                     breaks=c('Data','Seasonally Adjusted'))
```

The outlier seems to have the same effect regardless of whether it appears in the middle or towards the end of the time-series data.
