---
title: 'Data 624: Predictive Analytics HW:2'
author: 'Group 2: Alice Friedman, Kayleah Griffen, Michael Ippolito, Josh Iden'
date: "2023-06-19"
output:
  word_document:
    toc: yes
  pdf_document:
    dev: cairo_pdf
    toc: yes
  html_document:
    theme: yeti
    highlight: tango
    toc: yes
    toc_float: yes
always_allow_html: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo=TRUE, fig.width=9, fig.height=6, cache = T)
library(tidyverse)
library(kableExtra)
library(DMwR2)
library(gridExtra)
library(AppliedPredictiveModeling)
library(e1071)
library(caret)

# Load data
data(ChemicalManufacturingProcess)


library(reticulate)
use_python("C:\\GFApps\\Anaconda\\V3\\python.exe")
# use_python("C:/Users/kayle/anaconda3/python.exe")

```

```{python}
import pandas as pd
import numpy as np
from numpy import arange

from scipy.stats import skew

import matplotlib.pyplot as plt
from matplotlib import colors
from matplotlib.ticker import PercentFormatter

from sklearn.impute import SimpleImputer
from sklearn.feature_selection import VarianceThreshold
from sklearn.preprocessing import RobustScaler
from sklearn.preprocessing import PowerTransformer
from sklearn.model_selection import train_test_split
from sklearn import model_selection
from sklearn.model_selection import RepeatedKFold
from sklearn.cross_decomposition import PLSRegression
from sklearn.metrics import mean_squared_error
from sklearn.model_selection import cross_val_score
from sklearn.model_selection import RepeatedKFold
from sklearn.linear_model import ElasticNetCV


r.ChemicalManufacturingProcess.shape

r.ChemicalManufacturingProcess.isna().sum()

# define the imputer
imputer = SimpleImputer(strategy='mean')

# define the pandas dataframe to be imputed
pdfimp = r.ChemicalManufacturingProcess

# perform imputation
imputer = imputer.fit(pdfimp)
pdfimp.iloc[:,:] = imputer.transform(r.ChemicalManufacturingProcess)

# check nas and shape
pdfimp.isna().sum()
pdfimp.shape



var_thr = VarianceThreshold(threshold = 0.02) #Removing both constant and quasi-constant
var_thr.fit(pdfimp)

concol = [column for column in pdfimp.columns 
          if column not in pdfimp.columns[var_thr.get_support()]]

for features in concol:
    print(features)



pdf0var = pdfimp.drop(concol,axis=1)


for i in range(1,pdf0var.shape[1]):
  # IQR
  # Calculate the upper and lower limits
  Q1 = pdf0var.iloc[:, i].quantile(0.25)
  Q3 = pdf0var.iloc[:, i].quantile(0.75)
  IQR = Q3 - Q1
  lower = Q1 - 1.5*IQR
  upper = Q3 + 1.5*IQR
   
  # Create arrays of Boolean values indicating the outlier rows
  upper_array = np.where(pdf0var.iloc[:, i]>upper)[0]
  lower_array = np.where(pdf0var.iloc[:, i]<lower)[0]
  
  total = upper_array.shape[0] + lower_array.shape[0]
  
  print(pdf0var.columns[i], total)


all_skewed = []

for i in range(1,pdf0var.shape[1]):
  # Calculate the skewness
  s = skew(pdf0var.iloc[:, i], axis=0, bias=True)
  if (s > 2.0 or s < -2.0):
    print(pdf0var.columns[i], s)
    all_skewed.append(i)
  
pdfx = pdf0var.iloc[:,1:]




robust_scaler = RobustScaler()
pdfscaled = robust_scaler.fit_transform(pdfx.to_numpy())
pdfscaled = pd.DataFrame(pdfscaled, columns=pdfx.columns)


# can run this to see the plots
# for i in range(1,pdfx.shape[1]):
#   fig, axs = plt.subplots(1, 2, sharey=True, tight_layout=True)
#   n_bins = 30
#   axs[0].hist(pdfx.iloc[:, i], bins=n_bins)
#   axs[1].hist(pdfscaled.iloc[:, i], bins=n_bins)
#   plt.show()
  

power_transfromer = PowerTransformer(method='yeo-johnson', standardize=False)
pdfyeo = power_transfromer.fit_transform(pdfscaled.to_numpy())
pdfyeo = pd.DataFrame(pdfyeo, columns=pdfscaled.columns)

# for i in range(1,pdfscaled.shape[1]):
#   fig, axs = plt.subplots(1, 2, sharey=True, tight_layout=True)
#   n_bins = 30
#   axs[0].hist(pdfscaled.iloc[:, i], bins=n_bins)
#   axs[1].hist(pdfyeo.iloc[:, i], bins=n_bins)
#   plt.show()


f = plt.figure(figsize=(15, 15))
plt.matshow(pdfyeo.corr(), fignum=f.number)
plt.xticks(range(pdfyeo.select_dtypes(['number']).shape[1]), pdfyeo.select_dtypes(['number']).columns, fontsize=8, rotation=90)
plt.yticks(range(pdfyeo.select_dtypes(['number']).shape[1]), pdfyeo.select_dtypes(['number']).columns, fontsize=8)
cb = plt.colorbar()
cb.ax.tick_params(labelsize=14)
plt.show()


x = pdfyeo
y = r.ChemicalManufacturingProcess["Yield"]
  
# using the train test split function
x_train, x_test, y_train, y_test = train_test_split(x,y
,random_state=104,
test_size=0.2, 
shuffle=True)


#define cross-validation method
cv = RepeatedKFold(n_splits=10, n_repeats=3, random_state=1)

mse = []
n = len(x)

# Calculate MSE with only the intercept
score = -1*model_selection.cross_val_score(PLSRegression(n_components=1),
           np.ones((n,1)), y, cv=cv, scoring='neg_mean_squared_error').mean()    
mse.append(score)

# Calculate MSE using cross-validation, adding one component at a time
for i in range(1, 10):
    pls = PLSRegression(n_components=i)
    score = -1*model_selection.cross_val_score(pls, x, y, cv=cv,
               scoring='neg_mean_squared_error').mean()
    mse.append(score)
    
#plot test MSE vs. number of components
plt.clf()
plt.plot(mse)
plt.xlabel('Number of PLS Components')
plt.ylabel('MSE')
plt.show()


#calculate RMSE
pls = PLSRegression(n_components=5)
pls.fit((x_train), y_train)

np.sqrt(mse[6])



alphas = [0.0001, 0.001, 0.01, 0.1, 0.3, 0.5, 0.7, 1]

elastic_cv=ElasticNetCV(alphas=alphas, cv=cv)
enet = elastic_cv.fit(x_train, y_train)
print(enet.alpha_)



np.sqrt(mean_squared_error(y_test, pls.predict((x_test))))

np.sqrt(mean_squared_error(y_test, enet.predict(x_test)))


def _calculate_vips(model):
  t = model.x_scores_
  w = model.x_weights_
  q = model.y_loadings_
  p, h = w.shape
  vips = np.zeros((p,))
  s = np.diag(np.matmul(np.matmul(np.matmul(t.T,t),q.T), q)).reshape(h, -1)
  total_s = np.sum(s)
  for i in arange(p):
    weight = np.array([ (w[i,j] / np.linalg.norm(w[:,j]))**2 for j in range(h) ])
    vips[i] = np.sqrt(p*(np.matmul(s.T, weight))/total_s)
  return vips

vimps = _calculate_vips(pls)

vimpsdf = pd.DataFrame({'Var' : pdfscaled.columns, 'VarImp' : vimps})

vimpsdf.sort_values(by = "VarImp", ascending = False)


f = plt.figure(figsize=(15, 15))
plt.matshow(pdf0var.corr(), fignum=f.number)
plt.xticks(range(pdf0var.select_dtypes(['number']).shape[1]), pdf0var.select_dtypes(['number']).columns, fontsize=8, rotation=90)
plt.yticks(range(pdf0var.select_dtypes(['number']).shape[1]), pdf0var.select_dtypes(['number']).columns, fontsize=8)
cb = plt.colorbar()
cb.ax.tick_params(labelsize=14)
plt.show()
```


### 7.2

Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data:

y = 10 sin(πx1x2) + 20(x3 − 0.5)2 + 10x4 + 5x5 + N(0, σ2)

where the x values are random variables uniformly distributed between [0, 1] (there are also 5 other non-informative variables also created in the simulation). The package mlbench contains a function called mlbench.friedman1 that simulates these data:

```{r}
library(mlbench)
set.seed(200)
trainingData <- mlbench.friedman1(200, sd = 1)
## We convert the 'x' data from a matrix to a data frame
## One reason is that this will give the columns names.
trainingData$x <- data.frame(trainingData$x)
## Look at the data using
featurePlot(trainingData$x, trainingData$y)
## or other methods.

## This creates a list with a vector 'y' and a matrix
## of predictors 'x'. Also simulate a large test set to
## estimate the true error rate with good precision:
testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x)

```

Tune several models on these data. For example:

```{r}
library(caret)
knnModel <- train(x = trainingData$x, 
                    y = trainingData$y, 
                    method = "knn",
                    preProc = c("center", "scale"),
                    tuneLength = 10)
knnModel

knnPred <- predict(knnModel, newdata = testData$x)
## The function 'postResample' can be used to get the test set
## perforamnce values
postResample(pred = knnPred, obs = testData$y)

```

Which models appear to give the best performance? Does MARS select the informative predictors (those named X1–X5)?


```{python}
import numpy as np
from sklearn.neighbors import KNeighborsRegressor
knn_model = KNeighborsRegressor(n_neighbors=17)

data_train = r.trainingData.items()
data_train_list = list(data_train)
data_train_np = np.array(data_train_list)

knn_model.fit(data_train_np[0][1], data_train_np[1][1])

from sklearn.metrics import mean_squared_error
from math import sqrt

data_test = r.testData.items()
data_test_list = list(data_test)
data_test_np = np.array(data_test_list)

train_preds = knn_model.predict(data_test_np[0][1])
mse = mean_squared_error(data_test_np[1][1], train_preds)
rmse = sqrt(mse)
rmse
```

### 7.5

Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.

(a) Which nonlinear regression model gives the optimal resampling and test  set performance?

(b) Which predictors are most important in the optimal nonlinear regression model? Do either the biological or process variables dominate the list? How do the top ten important predictors compare to the top ten predictors from the optimal linear model?

(c) Explore the relationships between the top predictors and the response for the predictors that are unique to the optimal nonlinear regression model. Do these plots reveal intuition about the biological or process predictors and their relationship with yield?

### 8.1

Recreate the simulated data from Exercise 7.2

```{r}
library(mlbench)
set.seed(200)
simulated <- mlbench.friedman1(200, sd = 1)
simulated <- cbind(simulated$x, simulated$y)
simulated <- as.data.frame(simulated)
colnames(simulated)[ncol(simulated)] <- "y"
```


(a) Fit a random forest model to all of the predictors, then estimate the variable importance scores:

```{r}
library(randomForest)
library(caret)
model1 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp1 <- varImp(model1, scale = FALSE)

```

Did the random forest model significantly use the uninformative predictors (V6 – V10)?

(b) Now add an additional predictor that is highly correlated with one of the informative predictors. For example:

```{r}
simulated$duplicate1 <- simulated$V1 + rnorm(200) * .1
cor(simulated$duplicate1, simulated$V1)
```

Fit another random forest model to these data. Did the importance score for V1 change? What happens when you add another predictor that is also highly correlated with V1?

(c) Use the cforest function in the party package to fit a random forest model using conditional inference trees. The party package function varimp can calculate predictor importance. The conditional argument of that function toggles between the traditional importance measure and the modified version described in Strobl et al. (2007). Do these importances show the same pattern as the traditional random forest model?

(d) Repeat this process with different tree models, such as boosted trees and Cubist. Does the same pattern occur?

### 8.2

Use a simulation to show tree bias with different granularities.


### 8.3

In stochastic gradient boosting the bagging fraction and learning rate will govern the construction of the trees as they are guided by the gradient. Although the optimal values of these parameters should be obtained through the tuning process, it is helpful to understand how the magnitudes of these parameters affect magnitudes of variable importance. Figure 8.24 provides the variable importance plots for boosting using two extreme values for the bagging fraction (0.1 and 0.9) and the learning rate (0.1 and 0.9) for the solubility data. The left-hand plot has both parameters set to 0.1, and the right-hand plot has both set to 0.9:

(a) Why does the model on the right focus its importance on just the first few of predictors, whereas the model on the left spreads importance across more predictors?

(b) Which model do you think would be more predictive of other samples?

(c) How would increasing interaction depth affect the slope of predictor importance for either model in Fig. 8.24?

### 8.7

Refer to Exercises 6.3 and 7.5 which describe a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several tree-based models:

(a) Which tree-based regression model gives the optimal resampling and test
set performance?

(b) Which predictors are most important in the optimal tree-based regression model? Do either the biological or process variables dominate the list? How do the top 10 important predictors compare to the top 10 predictors from the optimal linear and nonlinear models?

(c) Plot the optimal single tree with the distribution of yield in the terminal nodes. Does this view of the data provide additional knowledge about the biological or process predictors and their relationship with yield?

### Recommender Problem
