---
title: "ippolito_data624_wk5"
author: "Michael Ippolito"
date: "2023-06-19"
output:
  html_document:
    theme: yeti
    highlight: tango
    toc: yes
    toc_float: yes
  pdf_document:
    dev: cairo_pdf
    toc: yes
---

```{r setup, include=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo=TRUE, fig.width=9, fig.height=6)
library(tidyverse)
library(kableExtra)
library(mice)
library(caret)
library(corrplot)
library(ggcorrplot)
library(PerformanceAnalytics)
library(MASS)  # for stepwise regression
library(forecast)
library(earth)  # for mars modeling
library(kernlab)  # for svm modeling
library(DMwR2)
library(gridExtra)

# Set minimal theme
theme_set(theme_minimal())

```

### KJ 7.2

Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data:

$y\ =\ 10sin(πx_1x_2)\ +\ 20(x_3\ −\ 0.5)^2\ +\ 10x_4\ +\ 5x_5\ +\ N(0,\ σ^2)$

where the x values are random variables uniformly distributed between [0, 1] (there are also 5 other non-informative variables also created in the simulation). The package mlbench contains a function called mlbench.friedman1 that simulates these data:

```{r}

library(mlbench)
set.seed(200)
trainingData <- mlbench.friedman1(200, sd = 1)
## We convert the 'x' data from a matrix to a data frame
## One reason is that this will give the columns names.
trainingData$x <- data.frame(trainingData$x)
## Look at the data using
featurePlot(trainingData$x, trainingData$y)
## or other methods.

## This creates a list with a vector 'y' and a matrix
## of predictors 'x'. Also simulate a large test set to
## estimate the true error rate with good precision:
testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x)

```

Tune several models on these data. For example:

```{r}

library(caret)
knnModel <- train(x = trainingData$x, 
                    y = trainingData$y, 
                    method = "knn",
                    preProc = c("center", "scale"),
                    tuneLength = 10)
knnModel

knnPred <- predict(knnModel, newdata = testData$x)
## The function 'postResample' can be used to get the test set
## perforamnce values
postResample(pred = knnPred, obs = testData$y)

```

First, set up a data frame for the results and set the cross-validation parameters.

```{r}

# Results data frame
dfr2 <- data.frame(matrix(nrow=10, ncol=3))
colnames(dfr2) <- c('Model', 'Tuning.Parameters', 'RMSE')

# specify 10x cross-validation
ctrl <- trainControl(method='cv', number=10)

# Add the knn results
dfr2[1,] = data.frame(
    Model='knn', 
    Tuning.Parameters=paste0('k=', knnModel$bestTune[['k']]), 
    RMSE=min(knnModel$results[['RMSE']])
)

```

Train a MARS model.

```{r}

# MARS model
marsGrid <- expand.grid(.degree=1:2, .nprune=2:10)  # set tuning parameters
set.seed(77)
fitmars <- train(x=trainingData$x, y=trainingData$y, method='earth', tuneGrid=marsGrid, trControl=ctrl)
fitmars
dfr2[2,] = data.frame(
    Model='MARS', 
    Tuning.Parameters=paste0('degree=', fitmars$bestTune[['degree']], ', nprune=', fitmars$bestTune[['nprune']]), 
    RMSE=min(fitmars$results[['RMSE']])
)

```

Train an SVM model.

```{r}

# SVM model
set.seed(77)
fitsvm <- train(x=trainingData$x, y=trainingData$y, method='svmRadial', preProc=c('center', 'scale'), tuneLength=14, trControl=ctrl)
fitsvm
dfr2[3,] = data.frame(
    Model='SVM', 
    Tuning.Parameters=paste0('C=', fitsvm$bestTune[['C']], ', sigma=', round(fitsvm$bestTune[['sigma']], 3)), 
    RMSE=min(fitsvm$results[['RMSE']])
)

```

Train a neural network model.

```{r warning=F}

# nnet model using model averaging
nnetGrid <- expand.grid(.decay=c(0, 0.01, 0.1), .size=c(1:10), .bag=F)  # set tuning parameters
set.seed(77)
fitnnet <- train(x=trainingData$x, y=trainingData$y, method='avNNet', preProc=c('center', 'scale'), tunGrid=nnetGrid, trControl=ctrl,
             linout=T, trace=F, MaxNWts=10 * (ncol(trainingData$x) + 1) + 10 + 1, maxit=500)
fitnnet
dfr2[4,] = data.frame(
    Model='avNNet', 
    Tuning.Parameters=paste0('decay=', fitnnet$bestTune[['decay']], ', size=', fitnnet$bestTune[['size']], ', bag=False'), 
    RMSE=min(fitnnet$results[['RMSE']])
)

```

Summarize models.

```{r}

# Summarize models
dfr2 %>%
    filter(!is.na(RMSE)) %>%
    kbl(caption='Model summary') %>%
    kable_classic(full_width=F)

```

**Which models appear to give the best performance? Does MARS select the informative predictors (those named X1–X5)?**

Surprisingly, the second-degree MARS model pruned back to 10 parameters performed the best, with an RMSE of 1.39--significantly less than that of its closest competitor, SVM, with an RMSE of 1.85. The model-averaged neural network model performed similarly to SVM (RMSE=2.00). The K-nearest neighbor model performed poorly, with an RMSE well above the others (3.18). This is not surprising, as KNN models typically perform poorly when there are predictors that don't contribute significantly to the response; as stated in the problem description, in this data set there are five such predictors. The other models account for "noisy" predictors by pruning or applying a weight or decay factor.


### KJ 7.5

**Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.**

We'll load, preprocess, and split the data as before.

```{r}

# Load data
library(AppliedPredictiveModeling)
data(ChemicalManufacturingProcess)
dfchem <- ChemicalManufacturingProcess  # To avoid typing this many letters

# Impute missing values
#imp <- mice(dfchem, printFlag=F)
imp <- knnImputation(dfchem, k=3)
dfchem2 <- complete(imp)

# Set seed
set.seed(77)

# Corr plot
corr1 <- cor(dfchem2)
high_corr <- findCorrelation(corr1, cutoff=0.9, exact=T, verbose=F, names=F)

# Remove the highly correlated variables
dfchem3 <- dfchem2[,-high_corr]

# Remove NZV features
tmp_nzv <- nearZeroVar(dfchem3)
dfchem4 <- dfchem3[,-tmp_nzv]

# Split into train/test; createDataPartition generates indicies of the training set
train_indices <- createDataPartition(dfchem4$Yield, p=0.80, times=1, list=F)
dftrain <- dfchem4[train_indices,]
dftest <- dfchem4[-train_indices,]

# Separate outcome and predictors
trainx <- dftrain %>% dplyr::select(-Yield)
trainy <- dftrain$Yield
testx <- dftest %>% dplyr::select(-Yield)
testy <- dftest$Yield

# specify 10x cross-validation
ctrl <- trainControl(method='cv', number=10)

# Results data frame
dfr <- data.frame(matrix(nrow=10, ncol=3))
colnames(dfr) <- c('Model', 'Tuning.Parameters', 'Train.RMSE')

```

KNN model.

```{r}

# KNN model
set.seed(77)
fitknn2 <- train(x=trainx, y=trainy, method='knn', tuneLength=10, trControl=ctrl, preProc=c('center', 'scale'))
fitknn2
dfr[1,] = data.frame(
    Model='knn', 
    Tuning.Parameters=paste0('k=', fitknn2$bestTune[['k']]), 
    Train.RMSE=min(fitknn2$results[['RMSE']])
)


```

MARS model.

```{r}

# MARS model
marsGrid <- expand.grid(.degree=1:2, .nprune=2:10)  # set tuning parameters
set.seed(77)
fitmars2 <- train(x=trainx, y=trainy, method='earth', tuneGrid=marsGrid, trControl=ctrl)
fitmars2
dfr[2,] = data.frame(
    Model='MARS', 
    Tuning.Parameters=paste0('degree=', fitmars2$bestTune[['degree']], ', nprune=', fitmars2$bestTune[['nprune']]), 
    Train.RMSE=min(fitmars2$results[['RMSE']])
)

```

SVM model.

```{r}

# SVM model
set.seed(77)
fitsvm2 <- train(x=trainx, y=trainy, method='svmRadial', preProc=c('center', 'scale'), tuneLength=14, trControl=ctrl)
fitsvm2
dfr[3,] = data.frame(
    Model='SVM', 
    Tuning.Parameters=paste0('C=', fitsvm2$bestTune[['C']], ', sigma=', round(fitsvm2$bestTune[['sigma']], 3)), 
    Train.RMSE=min(fitsvm2$results[['RMSE']])
)

```

Neural network model.

```{r}

# nnet model using model averaging
nnetGrid <- expand.grid(.decay=c(0, 0.01, 0.1), .size=c(1:10), .bag=c(T, F))  # set tuning parameters
set.seed(77)
fitnnet2 <- train(x=trainx, y=trainy, method='avNNet', preProc=c('center', 'scale'), tunGrid=nnetGrid, trControl=ctrl,
             linout=T, trace=F, MaxNWts=10 * (ncol(trainx) + 1) + 10 + 1, maxit=500)
fitnnet2
dfr[4,] = data.frame(
    Model='avNNet', 
    Tuning.Parameters=paste0('decay=', fitnnet2$bestTune[['decay']], ', size=', fitnnet2$bestTune[['size']], ', bag=', fitnnet2$bestTune[['bag']]), 
    Train.RMSE=min(fitnnet2$results[['RMSE']])
)

```

Display model summary.

```{r}

# Model summary
dfr %>%
    filter(!is.na(Train.RMSE)) %>%
    kbl(caption='Model summary') %>%
    kable_classic(full_width=F)

```

#### (a)

**Which nonlinear regression model gives the optimal resampling and test set performance?**

```{r}

# Add column for test set RMSE to results data frame
dfr$Test.RMSE = NA

# Predict using test set

# Knn
predy <- predict(fitknn2, newdata=as.matrix(testx))
dfr[1, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# MARS
predy <- as.vector(predict(fitmars2, newdata=as.matrix(testx)))
dfr[2, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# SVM
predy <- predict(fitsvm2, newdata=as.matrix(testx))
dfr[3, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Nnet
predy <- predict(fitnnet2, newdata=as.matrix(testx))
dfr[4, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Model summary
dfr %>%
    filter(!is.na(Train.RMSE)) %>%
    kbl(caption='Model summary') %>%
    kable_classic(full_width=F)

```

The first-degree MARS model pruned to six parameters yielded the best RMSE for both training and test data (1.19 and 1.01, respectively).

#### (b)

**Which predictors are most important in the optimal nonlinear regression model? Do either the biological or process variables dominate the list? How do the top ten important predictors compare to the top ten predictors from the optimal linear model?**

```{r}

varImp(fitmars2)$importance %>%
    kbl(caption='Variable imporance - MARS model') %>%
    kable_classic(full_width=F)

```

As shown above, ManufacturingProcess variables 32, 09, 13, and 39 are the most important features. These are only variables in the manufacturing process rather than biological variables, which again is advantageous since they can be controlled to some extent while the biological variables cannot be. The top three variables (32, 09, and 13) are the same as those found to be most important in the linear models, although in a different order (the linear model had ManufacturingProcess09 first). The fact that both linear and non-linear modeling chose the same top three predictors is good confirmation of their importance.

#### (c)

**Explore the relationships between the top predictors and the response for the predictors that are unique to the optimal nonlinear regression model. Do these plots reveal intuition about the biological or process predictors and their relationship with yield?**

```{r fig.width=9, fig.height=6}

# Plot response vs top predictors
p1 <- dfchem2 %>% ggplot(aes(x=ManufacturingProcess32, y=Yield)) +
    geom_point() +
    geom_smooth(method=lm, formula=y ~ x, linetype=2, color='darkred', se=F)
p2 <- dfchem2 %>% ggplot(aes(x=ManufacturingProcess09, y=Yield)) +
    geom_point() +
    geom_smooth(method=lm, formula=y ~ x, linetype=2, color='darkred', se=F)
p3 <- dfchem2 %>% ggplot(aes(x=ManufacturingProcess13, y=Yield)) +
    geom_point() +
    geom_smooth(method=lm, formula=y ~ x, linetype=2, color='darkred', se=F)
p4 <- dfchem2 %>% ggplot(aes(x=ManufacturingProcess39, y=Yield)) +
    geom_point() +
    geom_smooth(method=lm, formula=y ~ x, linetype=2, color='darkred', se=F)
p5 <- dfchem2 %>% filter(ManufacturingProcess39 > 0) %>%
    ggplot(aes(x=ManufacturingProcess39, y=Yield)) +
    geom_point() +
    geom_smooth(method=lm, formula=y ~ x, linetype=2, color='darkred', se=F) +
    ggtitle('ManufacturingProcess39 (removed zero values)')
grid.arrange(p1, p2, p3, p4, p5, ncol=2)

```

As in the linear model, processes 9 and 32 exhibit a positive relationship with yield, while the relationship between yield and process 13 is inverse. Process 39 is interesting in that there are some zero values that may or may not be outliers or data-entry errors. Including the outliers, there is a slight upward trend. The last graph shows that, after removing the potential outliers, the trend becomes slightly inverse. To assess whether these outliers should have been removed prior to modeling, we would have to know more about the context in which the source data were collected and whether zero values are expected from this type of process.

### 8.1

Recreate the simulated data from Exercise 7.2

```{r}

library(mlbench)
set.seed(200)
simulated <- mlbench.friedman1(200, sd = 1)
simulated <- cbind(simulated$x, simulated$y)
simulated <- as.data.frame(simulated)
colnames(simulated)[ncol(simulated)] <- "y"

```

#### (a)

**Fit a random forest model to all of the predictors, then estimate the variable importance scores:**

```{r}

library(randomForest)
library(caret)
model1 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp1 <- varImp(model1, scale = FALSE)

```

**Did the random forest model significantly use the uninformative predictors (V6 – V10)?**

```{r}

# Display importance values
rfImp1 %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Relative importance of predictors') %>%
    kable_classic(full_width=F)

```

As shown, the uninformative predictors (V6-V10) have a significantly lower relative importance in the model, which is consistent with our earlier findings.

#### (b)

**Now add an additional predictor that is highly correlated with one of the informative predictors. For example:**

```{r}

simulated$duplicate1 <- simulated$V1 + rnorm(200) * .1
cor(simulated$duplicate1, simulated$V1)

```

**Fit another random forest model to these data. Did the importance score for V1 change? What happens when you add another predictor that is also highly correlated with V1?**

```{r}

# Fit random forest to the new data with the highly corrected predictors
model2 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp2 <- varImp(model2, scale = FALSE)

# Display results
rfImp2 %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Relative importance of predictors - with highly correlated variables') %>%
    kable_classic(full_width=F)

```

Adding the duplicate1 variable caused the relative importance of V1 to decrease, while that of duplicate1 factored prominently in terms of importance--just below V1. In effect, the two variables "shared" how important they were in the model. This is analogous to the two surface area variables in the solubility data discussed in the textbook (KJ p. 186, 202-203). If another variable that is highly correlated to V1 were added to the data set, we would expect a further reduction in the importances of V1 and duplicate1, effectively sharing the importance between all three variables.

```{r}

# Add another variable that is highly correlated with V1
simulated$duplicate2 <- simulated$V1 + rnorm(200) * .1
cor(simulated$duplicate2, simulated$V1)

# Fit random forest to the new data with the highly corrected predictors
model3 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp3 <- varImp(model3, scale = FALSE)

# Display results
rfImp3 %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Relative importance of predictors - with highly correlated variables') %>%
    kable_classic(full_width=F)

```

As shown in the table, the relative importances of V1 and duplicate1 decreased with the introduction of variable duplicate2. The sum of these three importances are roughly on par with the importance of V1 in the first model and the sum of v1 + duplicate1 in the second model. This is consistent with the way regression trees are sensitive to highly correlated variables.

#### (c)

**Use the cforest function in the party package to fit a random forest model using conditional inference trees. The party package function varimp can calculate predictor importance. The conditional argument of that function toggles between the traditional importance measure and the modified version described in Strobl et al. (2007). Do these importances show the same pattern as the traditional random forest model?**

```{r}

# Init results table
dfr <- data.frame(matrix(nrow=0, ncol=3))
colnames(dfr) <- c('Model', 'Variable', 'Rank')

# Fit conditional inference random forest model
library(party)
cimodel1 <- cforest(y ~ ., data=simulated)

# Display variable importances - without conditional inference
data.frame(Overall=varimp(cimodel1, conditional=F)) %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Relative importance of predictors - without conditional inference') %>%
    kable_classic(full_width=F)

# Results table
dfr <- rbind(dfr, data.frame(
             Model='Random forest (no conditional inference)', 
             Variable=rownames(tmpdf),
             Rank=rank(-tmpdf$Overall)))

# Display variable importances - with conditional inference
tmpdf <- data.frame(Overall=varimp(cimodel1, conditional=T))
tmpdf %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Relative importance of predictors - with conditional inference') %>%
    kable_classic(full_width=F)

# Results table
dfr <- rbind(dfr, data.frame(
             Model='Random forest (with conditional inference)', 
             Variable=rownames(tmpdf),
             Rank=rank(-tmpdf$Overall)))

```

Using conditional inference has the effect of lowering the importance of the two variables that are highly correlated with V1. Variable V1 itself remained third in the order of relative importance among other variables. It didn't increase in importance and, in fact, actually had a lower magnitude, but the overall effect was that the importances of the duplicate variables were reduced. This can be attributed to the fact that statistical inference tests are exhaustively conducted against all possible predictors and across all possible split points, thereby penalizing models with higher numbers of splits.

#### (d)

**Repeat this process with different tree models, such as boosted trees and Cubist. Does the same pattern occur?**

First try a basic CART model.

```{r}

# Basic CART
library(rpart)
set.seed(77)
fitcart1 <- rpart(y ~ ., data=simulated)
tmpdf <- varImp(fitcart1)
tmpdf %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Variable imporance - basic CART') %>%
    kable_classic(full_width=F)

# Results table
dfr <- rbind(dfr, data.frame(
             Model='Basic CART', 
             Variable=rownames(tmpdf),
             Rank=rank(-tmpdf$Overall)))

```

Try a CART model with conditional inference.

```{r}

# Basic CART with conditional inference
library(party)
set.seed(77)
fitcart2 <- ctree(y ~ ., data=simulated)
tmpdf <- data.frame(Overall=attributes(fitcart2)$tree$criterion$statistic)
tmpdf %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Variable imporance - CART with conditional inference') %>%
    kable_classic(full_width=F)

# Results table
dfr <- rbind(dfr, data.frame(
             Model='Basic CART with conditional inference', 
             Variable=rownames(tmpdf),
             Rank=rank(-tmpdf$Overall)))

```

Regression model tree.

```{r}

# Regression model tree
library(RWeka)
set.seed(77)
fitrmt1 <- M5P(y ~ ., data=simulated)
fitrmt1

```

Rule-based model tree.

```{r}

# Rule-based model tree
set.seed(77)
fitrmt2 <- M5Rules(y ~ ., data=simulated)
fitrmt2

```

Bagged tree.

```{r}

# Bagged tree
library(ipred)
set.seed(77)
fitbag1 <- bagging(y ~ ., data=simulated)
tmpdf <- varImp(fitbag1)
tmpdf %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Variable imporance - bagged tree') %>%
    kable_classic(full_width=F)

# Results table
dfr <- rbind(dfr, data.frame(
             Model='Bagged tree', 
             Variable=rownames(tmpdf),
             Rank=rank(-tmpdf$Overall)))

```

Boosted tree.

```{r}

# Boosted tree
library(gbm)
set.seed(77)
fitgbm1 <- gbm(y ~ ., data=simulated, distribution='gaussian', n.trees=1000)
tmpdf <- varImp(fitgbm1, numTrees=1000)
tmpdf %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Variable imporance - boosted tree (1000 iterations)') %>%
    kable_classic(full_width=F)

# Results table
dfr <- rbind(dfr, data.frame(
             Model='Boosted tree', 
             Variable=rownames(tmpdf),
             Rank=rank(-tmpdf$Overall)))

```

Cubist.

```{r}

# Cubist
library(Cubist)
set.seed(77)
fitcub1 <- cubist(x=simulated %>% dplyr::select(-y), y=simulated$y)
tmpdf <- varImp(fitcub1, numTrees=1000)
tmpdf %>%
    arrange(desc(Overall)) %>%
    kbl(caption='Variable imporance - cubist') %>%
    kable_classic(full_width=F)

# Results table
dfr <- rbind(dfr, data.frame(
             Model='Cubist', 
             Variable=rownames(tmpdf),
             Rank=rank(-tmpdf$Overall)))

```

Compare variable importance of the models.

```{r warning=F}

# Plot variable importance
dfr %>%
    filter(Variable %in% c('V1', 'duplicate1', 'duplicate2')) %>%
    ggplot(aes(x=factor(Variable, level=c('V1', 'duplicate1', 'duplicate2')), y=11-Rank, group=Model, color=Model, shape=Model)) +
    geom_point() +
    geom_line() +
    scale_y_discrete(name='Rank', limits=rev(factor(seq(1, 10)))) +
    xlab('Variable') +
    ggtitle('Comparison of variable importance of highly correlated variables')

```

As shown in the graph, most trees tend to split the variable importance among the three correlated predictors, although most downplayed the importance of the second duplicated variable. The boosted tree and basic CART with conditional inference models seemed to handle duplicated variables the best, while the bagged tree performed poorly in that respect.

### 8.2

**Use a simulation to show tree bias with different granularities.**

```{r}

# see p. 182

```

### 8.3

In stochastic gradient boosting the bagging fraction and learning rate will govern the construction of the trees as they are guided by the gradient. Although the optimal values of these parameters should be obtained through the tuning process, it is helpful to understand how the magnitudes of these parameters affect magnitudes of variable importance. Figure 8.24 provides the variable importance plots for boosting using two extreme values for the bagging fraction (0.1 and 0.9) and the learning rate (0.1 and 0.9) for the solubility data. The left-hand plot has both parameters set to 0.1, and the right-hand plot has both set to 0.9.

#### (a)

**Why does the model on the right focus its importance on just the first few of predictors, whereas the model on the left spreads importance across more predictors?**

If the learning rate is set to a low value, there is less of a penalty against larger coefficients; therefore, less influential predictors will remain small relative to more influential predictors. On the other hand, if the learning rate is high, the penalty against larger coefficients will be greater, resulting in reduced influence for these predictors. The idea behind increasing the learning rate is to try reducing overfitting. This results in fewer variables with greater importance--i.e., a more parsimonious model. The model on the left, with a lower learning rate, could be overfit, as there are more variables with greater importance (KJ pp. 206-207).

Likewise, as the bagging fraction increases, a greater proportion of the samples are used in each model; this results the same variables being used in many of the trees, in effect increasing the importance of those variables and causing the steeper gradient exhibited by the graph with the higher bagging fraction.

#### (b)

**Which model do you think would be more predictive of other samples?**

The model on the right would most likely yield better predictive performance than the model on the left. This is because the left model would prone to overfitting and, additionally, uses less of the training data at each iteration to build the model. In general, models that are overfit and which use a dataset that underrepresents the larger population of the data have poorer predictive performances.

#### (c)

**How would increasing interaction depth affect the slope of predictor importance for either model in Fig. 8.24?**

We would expect the slope of the predictor importance to be steeper if the interaction depth were increased. Because trees with more levels will have a greater number of weak predictors than less complex trees, the difference between strong and weak predictors will be more dramatic, resulting in a higher gradient between the two.

### 8.7

**Refer to Exercises 6.3 and 7.5 which describe a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several tree-based models.**

We'll load, preprocess, and split the data as before.

```{r}

# Load data
library(AppliedPredictiveModeling)
data(ChemicalManufacturingProcess)
dfchem <- ChemicalManufacturingProcess  # To avoid typing this many letters

# Impute missing values
#imp <- mice(dfchem, printFlag=F)
imp <- knnImputation(dfchem, k=3)
dfchem2 <- complete(imp)

# Set seed
set.seed(77)

# Corr plot
corr1 <- cor(dfchem2)
high_corr <- findCorrelation(corr1, cutoff=0.9, exact=T, verbose=F, names=F)

# Remove the highly correlated variables
dfchem3 <- dfchem2[,-high_corr]

# Remove NZV features
tmp_nzv <- nearZeroVar(dfchem3)
dfchem4 <- dfchem3[,-tmp_nzv]

# Split into train/test; createDataPartition generates indicies of the training set
train_indices <- createDataPartition(dfchem4$Yield, p=0.80, times=1, list=F)
dftrain <- dfchem4[train_indices,]
dftest <- dfchem4[-train_indices,]

# Separate outcome and predictors
trainx <- dftrain %>% dplyr::select(-Yield)
trainy <- dftrain$Yield
testx <- dftest %>% dplyr::select(-Yield)
testy <- dftest$Yield

# specify 10x cross-validation
ctrl <- trainControl(method='cv', number=10)

# Results data frame
dfr <- data.frame(matrix(nrow=10, ncol=3))
colnames(dfr) <- c('Model', 'Tuning.Parameters', 'Train.RMSE')

```

Basic CART model tuned using complexity parameter.

```{r}

# Basic CART model - tuned using complexity parameter
set.seed(77)
fitcart1 <- train(trainx, trainy, method='rpart', tuneLength=10, trControl=ctrl)
fitcart1
dfr[1,] = data.frame(
    Model='Basic CART (tuned w/complexity parameter)', 
    Tuning.Parameters=paste0('cp=', fitcart1$bestTune[['cp']]), 
    Train.RMSE=min(fitcart1$results[['RMSE']])
)


```

Basic CART model tuned using node depth.

```{r}

# Basic CART model - tuned using complexity parameter
set.seed(77)
fitcart2 <- train(trainx, trainy, method='rpart2', tuneLength=10, trControl=ctrl)
fitcart2
dfr[2,] = data.frame(
    Model='Basic CART (tuned w/node depth)', 
    Tuning.Parameters=paste0('maxdepth=', fitcart2$bestTune[['maxdepth']]), 
    Train.RMSE=min(fitcart2$results[['RMSE']])
)

```

Bagged CART.

```{r}

# Bagged CART
set.seed(77)
fitbag1 <- train(trainx, trainy, method='treebag', trControl=ctrl)
fitbag1
dfr[3,] = data.frame(
    Model='Bagged CART', 
    Tuning.Parameters='', 
    Train.RMSE=min(fitbag1$results[['RMSE']])
)

```

Random forest.

```{r}

# Random forest
set.seed(77)
fitrf1 <- train(trainx, trainy, method='rf', tuneLength=10, trControl=ctrl)
fitrf1
dfr[4,] = data.frame(
    Model='Random forest', 
    Tuning.Parameters=paste0('mtry=', fitrf1$bestTune[['mtry']]), 
    Train.RMSE=min(fitrf1$results[['RMSE']])
)

```

Stochastic gradient boosting.

```{r message=FALSE}

# Had to set message=FALSE to prevent thousands of trace messages

# Stochastic gradient boosting
gbmGrid <- expand.grid(.interaction.depth=seq(1, 7, by=2),
                       .n.trees=seq(100, 1000, by=50),
                       .shrinkage=c(0.01, 0.1),
                       .n.minobsinnode=10)
set.seed(77)
fitgbm1 <- train(trainx, trainy, method='gbm', tuneGrid=gbmGrid, trControl=ctrl)

```

```{r}

fitgbm1
dfr[5,] = data.frame(
    Model='Stochastic gradient boosting', 
    Tuning.Parameters=paste0('interaction.depth=', fitgbm1$bestTune[['interaction.depth']], 
                             ', n.trees=', fitgbm1$bestTune[['n.trees']], 
                             ', shrinkage=', fitgbm1$bestTune[['shrinkage']],
                             ', n.minobsinnode=10'), 
    Train.RMSE=min(fitgbm1$results[['RMSE']])
)

```

Cubist.

```{r}

# Cubist
cubGrid <- expand.grid(.committees=c(seq(1, 10), seq(20, 100, by=10)), .neighbors=c(0, 1, 5, 9))
set.seed(77)
fitcub1 <- train(trainx, trainy, method='cubist', tuneGrid=cubGrid, trControl=ctrl)
fitcub1
dfr[6,] = data.frame(
    Model='Cubist', 
    Tuning.Parameters=paste0('committees=', fitcub1$bestTune[['committees']], 
                             ', neighbors=', fitcub1$bestTune[['neighbors']]), 
    Train.RMSE=min(fitcub1$results[['RMSE']])
)

```

####(a)

**Which tree-based regression model gives the optimal resampling and test set performance?**

First generate predictions on the test set for each model, then show the summary of results.

```{r}

# Add column for test set RMSE to results data frame
dfr$Test.RMSE = NA

# Predict using test set

# Basic CART model - with complexit parameter
predy <- predict(fitcart1, newdata=as.matrix(testx))
dfr[1, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Basic CART model - with node depth
predy <- predict(fitcart2, newdata=as.matrix(testx))
dfr[2, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Bagged CART
predy <- predict(fitbag1, newdata=as.matrix(testx))
dfr[3, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Random forest
predy <- predict(fitrf1, newdata=as.matrix(testx))
dfr[4, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Stochastic gradient boosting
predy <- predict(fitgbm1, newdata=as.matrix(testx))
dfr[5, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Cubist
predy <- predict(fitcub1, newdata=as.matrix(testx))
dfr[6, 'Test.RMSE'] <- accuracy(predy, testy)[[2]]

# Model summary
dfr %>%
    filter(!is.na(Train.RMSE)) %>%
    mutate(Train.RMSE=round(Train.RMSE, 3)) %>%
    arrange(Train.RMSE) %>%
    kbl(caption='Model summary') %>%
    kable_classic(full_width=F)

# Graph results
dfr %>%
    filter(!is.na(Train.RMSE)) %>%
    ggplot() +
    geom_point(aes(x=reorder(Model, Train.RMSE), y=Train.RMSE, color='10xCV Training Set'), shape=1) +
    geom_point(aes(x=Model, y=Test.RMSE, color='Validation Set'), shape=2) +
    coord_flip() +
    ylab('RMSE') +
    ggtitle('Model summary')

```

The stochastic gradient boosting model

#### (b)

**Which predictors are most important in the optimal tree-based regression model? Do either the biological or process variables dominate the list? How do the top 10 important predictors compare to the top 10 predictors from the optimal linear and nonlinear models?**

```{r}

varImp(fitcub1)$importance %>%
    arrange(desc(Overall)) %>%
    head(n=10) %>%
    kbl(caption='Top 10 predictors of best-performing model (Cubist)') %>%
    kable_classic(full_width=F)

plot(varImp(fitcub1), top=10, main="Top 10 predictors of best-performing model (Cubist)")

```

All three classes of models (linear, non-linear, and tree-based) placed manufacturing process (MP) 32 as its most important predictor. MP 09 factored very high in all three as well (second in the linear and non-linear and fourth in the tree-based models), as did MP 13 (third in the linear and non-linear models and fifth in the tree-based models). Interestingly, the tree-based models ranked biologial material (BM) 06 as the second-most important predictor, while neither the linear nor non-linearl models included any BM predictors. This may be due to the tendency of trees to boost the prediction importance of variables that are related in a collinear fashion.

#### (c)

**Plot the optimal single tree with the distribution of yield in the terminal nodes. Does this view of the data provide additional knowledge about the biological or process predictors and their relationship with yield?**

```{r fig.width=10, fig.height=8}

# Try plotting cubist trees
if (1 == 0) {
    library(ggraph)
    library(igraph)
    fitcub1$finalModel$splits
    g <- graph_from_data_frame(fitcub1$finalModel$splits)
    g
    ggraph(g, layout='dendrogram', circular=F) + 
        geom_edge_diagonal() +
        geom_node_point() +
        theme_void() + 
        geom_node_label(aes(label=name))
}

# Give up and plot single tree; first fit a single-tree CART model
fitcart3 <- rpart(Yield ~ ., data=dfchem3)

# Plot
library(rattle)
fancyRpartPlot(fitcart3, palette='YlGnBu', caption="Single CART")

```

The single-CART tree view confirms that MP 32 is the most important predictor, as it was with the best-performing tree, linear, and non-linear models. It is interesting to note that the single tree is split on multiple biological materials, while the optimal Cubist model only lists a single one (BM 06) which doesn't factor prominently in the single tree (it is on the fifth level). So while this may provide some information about the relationship between yield and the predictors, it doesn't necessarily provide the **best** information.
