---
title: "ippolito_data624_wk6"
author: "Michael Ippolito"
date: "2023-07-04"
output: html_document
---

output:
  html_document:
    theme: yeti
    highlight: tango
    toc: yes
    toc_float: yes
  pdf_document:
    dev: cairo_pdf
    toc: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo=TRUE, fig.width=9, fig.height=6)
library(tidyverse)
library(kableExtra)

# Set minimal theme
theme_set(theme_minimal())

```

**Market basket analysis**

Imagine 10000 receipts sitting on your table. Each receipt represents a transaction with items that were purchased. The receipt is a representation of stuff that went into a customer’s basket – and therefore ‘Market Basket Analysis’.

That is exactly what the Groceries Data Set contains: a collection of receipts with each line representing 1 receipt and the items purchased. Each line is called a transaction and each column in a row represents an item.

Here is the dataset = GroceryDataSet.csv  (comma separated file)

You assignment is to use R to mine the data for association rules.  You should report support, confidence and lift and your top 10 rules by lift.  Turn in as you would the other problems from HA and KJ.  You should packaged this with your HW #2 set.

NOTE:  Bonus - Use a Python Library in addition to R and get a Bonus.

### Load data

```{r}

# Load data from github
dfmkt_raw <- read.csv('https://raw.githubusercontent.com/klgriffen96/summer23_data624/main/hw_2/GroceryDataSet.csv', header=F)

```

### Data preparation

```{r}

# Add column for transaction number and reorder so that it is first
dfmkt <- dfmkt_raw
dfmkt$trans.num <- row.names(dfmkt_raw)
dfmkt <- dfmkt %>%
    dplyr::select(trans.num, 1:ncol(dfmkt) - 1)

# Gather each item into its own row (i.e. make df into "long" format)
dfmkt <- gather(dfmkt, key='item.num', value='item.name', 2:ncol(dfmkt)) %>% arrange(trans.num, item.num)

# Remove items that are blank
dfmkt <- dfmkt %>%
    filter(item.name != '')

# Create a variable to indicate the presence of this item for this transaction
dfmkt$val <- 1

# Spread to "wide" format; this will fill in "1" into columns that have that item
# and NAs into columns that don't have that item
dfmkt2 <- spread(dfmkt, key='item.name', value='val', fill=NA)

# Replace the NAs with zeros
dfmkt2[is.na(dfmkt2)] <- 0

# Combine rows by trnasaction num
dfmkt2 <- dfmkt2 %>%
    group_by(trans.num) %>%
    summarize(across(c(-item.num), list(sum))) %>%
    ungroup()

```

### EDA

```{r}

# Show unique item count
print(paste0('There are ', length(unique(dfmkt$item.name)), ' unique items.'))

# Get item counts
dfgrp <- dfmkt %>%
    group_by(item.name) %>%
    summarize(ct=n()) %>%
    ungroup() %>%
    arrange(desc(ct))

# Most frequently purchased items
#dfgrp %>%
#    head(10) %>%
#    kbl(caption='10 Most purchased items') %>%
#    kable_classic(full_width=F)

# Most frequently purchased items
dfgrp %>%
    arrange(desc(ct)) %>%
    head(10) %>%
    ggplot(aes(x=reorder(item.name, ct), y=ct)) +
    geom_bar(stat='identity', width=0.5) +
    geom_text(aes(label=ct), hjust=-0.1) +
    coord_flip() +
    xlab('Item') +
    ylab('Count') +
    ggtitle('Most puchased items')

# Infrequently purchased items
#dfgrp %>%
#    arrange(ct) %>%
#    head(10) %>%
#    kbl(caption='10 Least purchased items') %>%
#    kable_classic(full_width=F)

# Infrequently purchased items
dfgrp %>%
    arrange(ct) %>%
    head(10) %>%
    ggplot(aes(x=reorder(item.name, desc(ct)), y=ct)) +
    geom_bar(stat='identity', width=0.5) +
    geom_text(aes(label=ct), hjust=-0.1) +
    coord_flip() +
    xlab('Item') +
    ylab('Count') +
    ggtitle('Least puchased items')

# Numbers of items purchased per transaction
data.frame(Items=rowSums(dfmkt2[,2:ncol(dfmkt2)])) %>%
    ggplot() +
    geom_histogram(aes(x=Items), bins=30) +
    xlab('Items purchased per transaction') +
    ylab('') +
    ggtitle('Number of items purchased per transaction')

```

### Association rules

```{r}

# Load required packages
library(arules)
library(arulesViz)

# Create transactional database from data set
trans <- read.transactions('https://raw.githubusercontent.com/klgriffen96/summer23_data624/main/hw_2/GroceryDataSet.csv', 
    format='basket', rm.duplicates=F, sep=',')

# Create associate rules
arules <- apriori(trans, parameter=list(supp = 0.01, conf=0.5, maxlen=10))

# Show top 10 rules
inspect(rev(arules[1:10])) %>%
    kbl(caption='Top 10 association rules') %>%
    kable_classic(full_width=F)

```
