---
title: 'Data 624: Predictive Analytics HW 2'
author: "Kayleah Griffen"
date: "5/21/2023"
output:
  word_document: default
  html_document:
    df_print: paged
always_allow_html: true
---

## Introduction

This homework assignment includes problems from:

Kuhn & Johnson. "Applied Predictive Modeling"

This accompanies readings from KJ Chapters 6, 7, and 8.

Additionally a recommender problem was assigned by the professor.

## Homework Solutions

```{r}
library(pls)
library(caret)
library(corrplot)
library(caret)
library(tidyverse)
library(psych)
library(imputeTS)
library(forecast)
library(AppliedPredictiveModeling)
library(DMwR2)

```

### 6.3 

A chemical manufacturing process for a pharmaceutical product was discussed in Sect. 1.4. In this problem, the objective is to understand the relationship between biological measurements of the raw materials (predictors), measurements of the manufacturing process (predictors), and the response of product yield. Biological predictors cannot be changed but can be used to assess the quality of the raw material before processing. On the other hand, manufacturing process predictors can be changed in the manufacturing process. Improving product yield by 1 % will boost revenue by approximately
one hundred thousand dollars per batch:

(a) Start R and use these commands to load the data:

```{r}
# book had lowercase C - need uppercase https://rdrr.io/rforge/AppliedPredictiveModeling/man/ChemicalManufacturingProcess.html
data("ChemicalManufacturingProcess")
```

The matrix processPredictors contains the 57 predictors (12 describing the input biological material and 45 describing the process predictors) for the 176 manufacturing runs. yield contains the percent yield for each run.

(b) A small percentage of cells in the predictor set contain missing values. Use an imputation function to fill in these missing values (e.g., see Sect. 3.8).

```{r}
# impute 
# https://www.rdocumentation.org/packages/DMwR2/versions/0.0.2/topics/knnImputation
cmp_i <- knnImputation(ChemicalManufacturingProcess,k=3)
```

(c) Split the data into a training and a test set, pre-process the data, and
tune a model of your choice from this chapter. What is the optimal value
of the performance metric?


First slit up the data.


```{r}
set.seed("09041996")

## Split training and test
i <- createDataPartition(cmp_i$Yield,
                                    p = 0.8,
                                    list = FALSE,
                                    times = 1)

x <- cmp_i[,colnames(cmp_i)[colnames(cmp_i) != 'Yield']]
y <- cmp_i[,colnames(cmp_i)[colnames(cmp_i) == 'Yield']]

data.train.x <- x[i,]
data.train.y <- y[i]
data.test.x <- x[-i,]
data.test.y <- y[-i]
```

Next check the distributions and check for near zero variance and outliers.

```{r, fig.height = 3, fig.width = 3, fig.align = "center"}
## Pre-process data

# take a look at the distributions of predictors
# for (i in 1:length(data.train.x)){
#   p <- data.train.x |> ggplot(aes(x = data.train.x[[i]])) + geom_histogram()
#   print(p)
# }
# 
# nzv <- nearZeroVar(data.train.x)
# trainData <- trainData[-nzv]
# paste0("Inds near zero variance: ", colnames(data.train.x[nzv]))

# check for outliers
# for (i in 1:length(data.train.x)){
#   p <- data.train.x |> ggplot(aes(x = data.train.x[[i]])) + 
#     geom_boxplot() + xlab(colnames(data.train.x[i]))
#   print(p)
# }


# for (i in 1:length(data.train.x)){
#   data <- data.train.x[[i]]
#   quartiles <- quantile(data, probs=c(.25, .75), na.rm = FALSE)
#   IQR <- IQR(data)
#   Lower <- quartiles[1] - 1.5*IQR
#   Upper <- quartiles[2] + 1.5*IQR 
#   outliers <- which(data < Lower | data > Upper)
#   data[outliers] <- NA
#   print(length(outliers))
#   data.train.x[[i]] <- data
# }

# for (i in 1:length(data.train.x)){
#   p <- data.train.x |> ggplot(aes(x = data.train.x[[i]])) + 
#     geom_boxplot() + xlab(colnames(data.train.x[i]))
#   print(p)
# }
```

**NOTE some variables had several outliers

Next check correlations between variables.

```{r, fig.height = 10, fig.width = 10, fig.align = "center"}

# take a look at correlations

data.train.x.corr <- cor(data.train.x)

# Corr chart
corrplot(data.train.x.corr, order = "AOE", type = "lower", diag = FALSE)

# Find correlations past cutoff
print(paste0("Candidate for removal due to high correlation: ", findCorrelation(data.train.x.corr, cutoff=0.8, exact=T, verbose=T, names=T)))

```

Last check skewness.

```{r}
# check skewness
# df<- describe(data.train.x) %>% select(skew)
# inds_for_log <- which(abs(df$skew) > 1)
# paste0("Inds skewed: ", colnames(data.train.x[inds_for_log]))
```

Try using preProcess for preProcessing.

```{r}
trans <- preProcess(data.train.x, method = c("center", "scale", "YeoJohnson", "nzv"))
data.train.x.trans <- predict(trans, data.train.x)
```

Now that the data is preprocessed, try a model.

```{r}

lmFitAllPredictors <- lm(data.train.y ~ ., data = data.train.x.trans)
summary(lmFitAllPredictors)# Model of choice 


```

```{r}
ctrl <- trainControl(method = "cv", number = 10)
lmFit1 <- train(x = data.train.x, 
                y = data.train.y, method = "lm", trControl = ctrl)
lmFit1$results
```


(d) Predict the response for the test set. What is the value of the performance
metric and how does this compare with the resampled performance metric
on the training set?

```{r}
trans <- preProcess(data.test.x, method = c("center", "scale", "YeoJohnson", "nzv"))
data.test.x.trans <- predict(trans, data.test.x)

lm.pred = predict(lmFit1, data.test.x)

accuracy(lm.pred, data.test.y)
```


```{r}
# rlmPCA <- train(data.train.x.trans, 
#                 data.train.y, 
#                 method = "rlm", preProcess = "pca", Control = ctrl)
# 
# rlmPCA

```

(e) Which predictors are most important in the model you have trained? Do
either the biological or process predictors dominate the list?

```{r}
df_imp <- varImp(lmFit1)
df_imp
```

(f) Explore the relationships between each of the top predictors and the response. How could this information be helpful in improving yield in future
runs of the manufacturing process?

```{r}
inds_imp <- which(df_imp$importance$Overall > 40)
rnames_imp <- rownames(df_imp$importance)[inds_imp]
rnames_imp <- append(rnames_imp, "Yield")

imp_cor <- cor(cmp_i[,c(rnames_imp)])

# Corr chart
corrplot(cor(imp_cor), method="number", order = "AOE", type = "lower", diag = FALSE, na.label = "?")

```

### 7.2

Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data:

y = 10 sin(πx1x2) + 20(x3 − 0.5)2 + 10x4 + 5x5 + N(0, σ2)

where the x values are random variables uniformly distributed between [0, 1] (there are also 5 other non-informative variables also created in the simulation). The package mlbench contains a function called mlbench.friedman1 that simulates these data:

```{r}
library(mlbench)
set.seed(200)
trainingData <- mlbench.friedman1(200, sd = 1)
## We convert the 'x' data from a matrix to a data frame
## One reason is that this will give the columns names.
trainingData$x <- data.frame(trainingData$x)
## Look at the data using
featurePlot(trainingData$x, trainingData$y)
## or other methods.

## This creates a list with a vector 'y' and a matrix
## of predictors 'x'. Also simulate a large test set to
## estimate the true error rate with good precision:
testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x)

```

Tune several models on these data. For example:

```{r}
library(caret)
knnModel <- train(x = trainingData$x, 
                    y = trainingData$y, 
                    method = "knn",
                    preProc = c("center", "scale"),
                    tuneLength = 10)
knnModel

knnPred <- predict(knnModel, newdata = testData$x)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.knn <- postResample(pred = knnPred, obs = testData$y)

```


Next try neural network.

```{r}

# remove predictors so absolute pairwise correlation is less than 0.75

tooHigh <- findCorrelation(cor(trainingData$x), cutoff = .75)

# no correlations greater than 0.75! 

library(doParallel)
# Find out how many cores are available: 
detectCores()
## [1] 16
# Create cluster with desired number of cores: 
cl <- makeCluster(16)
# Register cluster: 
registerDoParallel(cl)
# Find out how many cores are being used
getDoParWorkers()

nnetGrid <- expand.grid(.decay = c(0, 0.01, .1), .size = c(1:10), .bag = FALSE)

nnetModel <- train(x = trainingData$x, 
                    y = trainingData$y,
                   method = "avNNet",
                   tuneGrid = nnetGrid,
                   trControl = trainControl(method = "cv"),
                   preProc = c("center", "scale"),
                   linout = TRUE,
                   trace = FALSE,
                   MaxNWts = 10 * (ncol(trainingData$x) + 1) + 10 + 1,
                   maxit = 500,
                   allowParallel = TRUE)

stopCluster(cl)
registerDoSEQ()

# nnetModel

nnetPredict <- predict(nnetModel, newdata = testData$x)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.nnet <- postResample(pred = nnetPredict, obs = testData$y)

```

Next try MARS. 

```{r}

marsGrid <- expand.grid(.degree = 1:2, .nprune = 2:38)

MARSModel <- train(x = trainingData$x, 
                    y = trainingData$y, 
                    method = "earth",
                    tuneGrid = marsGrid,
                    trControl = trainControl(method = "cv"))

MARSModel

MARSPred <- predict(MARSModel, newdata = testData$x)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.MARS <- postResample(pred = MARSPred, obs = testData$y)

varImp(MARSModel)

```

Next try SVM.

```{r}
svmModel <- train(x = trainingData$x, 
                    y = trainingData$y, 
                    method = "svmRadial",
                    preProc = c("center", "scale"),
                  tuneLength = 14,
                  trControl = trainControl(method = "cv"))
svmModel

svmPred <- predict(svmModel, newdata = testData$x)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.svm <- postResample(pred = svmPred, obs = testData$y)

```


Which models appear to give the best performance? Does MARS select the informative predictors (those named X1–X5)?

MARS did select the informative predictor, but one thing to note is it eliminated X3.

```{r}
p.knn
p.nnet
p.MARS
p.svm
```

The best rquared was from the MARS model along with the best RMSE and MAE.

### 7.5

Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.

```{r}
data("ChemicalManufacturingProcess")
cmp_i <- knnImputation(ChemicalManufacturingProcess,k=3)

cmp_i2 <- cmp_i |> select(-Yield)

trans <- preProcess(cmp_i2, method = c("center", "scale", "YeoJohnson", "nzv"))
cmp_i2 <- predict(trans, cmp_i2)

cmp_i2$Yield <- cmp_i$Yield

cmp_i <- cmp_i2

i <- createDataPartition(cmp_i$Yield,
                                    p = 0.8,
                                    list = FALSE,
                                    times = 1)

x <- cmp_i[,colnames(cmp_i)[colnames(cmp_i) != 'Yield']]
y <- cmp_i[,colnames(cmp_i)[colnames(cmp_i) == 'Yield']]

data.train.x <- x[i,]
data.train.y <- y[i]
data.test.x <- x[-i,]
data.test.y <- y[-i]
```

Tune several models on these data. For example:

```{r}
library(caret)
knnModel <- train(x = data.train.x, 
                    y = data.train.y, 
                    method = "knn",
                    tuneLength = 10)
knnModel

knnPred <- predict(knnModel, newdata = data.test.x)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.knn <- postResample(pred = knnPred, obs = data.test.y)

```


Next try neural network.

```{r}

# remove predictors so absolute pairwise correlation is less than 0.75

tooHigh <- findCorrelation(cor(data.train.x), cutoff = .75)


trainXnnet <- data.train.x[, -tooHigh]
testXnnet <- data.test.x[, -tooHigh]

# Find out how many cores are available: 
detectCores()
## [1] 16
# Create cluster with desired number of cores: 
cl <- makeCluster(16)
# Register cluster: 
registerDoParallel(cl)
# Find out how many cores are being used
getDoParWorkers()

nnetGrid <- expand.grid(.decay = c(0, 0.01, .1), .size = c(1:10), .bag = FALSE)

nnetModel <- train(x = trainXnnet, 
                    y = data.train.y,
                   method = "avNNet",
                   tuneGrid = nnetGrid,
                   trControl = trainControl(method = "cv"),
                   preProc = c("center", "scale"),
                   linout = TRUE,
                   trace = FALSE,
                   MaxNWts = 10 * (ncol(trainXnnet) + 1) + 10 + 1,
                   maxit = 500,
                   allowParallel = TRUE)

stopCluster(cl)
registerDoSEQ()

# nnetModel

nnetPredict <- predict(nnetModel, newdata = testXnnet)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.nnet <- postResample(pred = nnetPredict, obs = data.test.y)

```

Next try MARS. 

```{r}

marsGrid <- expand.grid(.degree = 1:2, .nprune = 2:38)


MARSModel <- train(x = data.train.x, 
                    y = data.train.y, 
                    method = "earth",
                    tuneGrid = marsGrid,
                    trControl = trainControl(method = "cv"))


MARSModel

MARSPred <- predict(MARSModel, newdata = data.test.x)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.MARS <- postResample(pred = MARSPred, obs = data.test.y)


```

Next try SVM.

```{r}
svmModel <- train(x = data.train.x, 
                    y = data.train.y, 
                    method = "svmRadial",
                  tuneLength = 14,
                  trControl = trainControl(method = "cv"))
svmModel

svmPred <- predict(svmModel, newdata = data.test.x)
## The function 'postResample' can be used to get the test set
## perforamnce values
p.svm <- postResample(pred = svmPred, obs = data.test.y)

```


```{r}
p.knn
p.nnet
p.MARS
p.svm
```


(a) Which nonlinear regression model gives the optimal resampling and test  set performance?

The optimal resampling and test performance was from the MARS model.


(b) Which predictors are most important in the optimal nonlinear regression model? Do either the biological or process variables dominate the list? How do the top ten important predictors compare to the top ten predictors from the optimal linear model?

```{r}
varImp(MARSModel)

```


Similar to the optimal linear model - process variables dominate the variable importance list. Similar to the linear model - ManufacturingProcess32 is registered as the most important. The most important biological process is different. 

(c) Explore the relationships between the top predictors and the response for the predictors that are unique to the optimal nonlinear regression model. Do these plots reveal intuition about the biological or process predictors and their relationship with yield?

```{r}
df_imp <- varImp(MARSModel)
inds_imp <- which(df_imp$importance$Overall > 0)
rnames_imp <- rownames(df_imp$importance)[inds_imp]
rnames_imp <- append(rnames_imp, "Yield")

imp_cor <- cor(cmp_i[,c(rnames_imp)])

# Corr chart
corrplot(cor(imp_cor), method="number", order = "AOE", type = "lower", diag = FALSE, na.label = "?")

```

### 8.1

Recreate the simulated data from Exercise 7.2

```{r}
library(mlbench)
set.seed(200)
simulated <- mlbench.friedman1(200, sd = 1)
simulated <- cbind(simulated$x, simulated$y)
simulated <- as.data.frame(simulated)
colnames(simulated)[ncol(simulated)] <- "y"
```


(a) Fit a random forest model to all of the predictors, then estimate the
variable importance scores:

```{r}
library(randomForest)
library(caret)
model1 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp1 <- varImp(model1, scale = FALSE)
rfImp1
```

Did the random forest model significantly use the uninformative predictors (V6 – V10)?

The random forest model did not signifigantly use the uninformative predictors - all of their importance was less than a magnitude of 0.2.

(b) Now add an additional predictor that is highly correlated with one of the
informative predictors. For example:

```{r}
simulated$duplicate1 <- simulated$V1 + rnorm(200) * .1
cor(simulated$duplicate1, simulated$V1)
```

Fit another random forest model to these data. Did the importance score for V1 change? 

```{r}
model2 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp2 <- varImp(model2, scale = FALSE)
rfImp2
```

The variable importance of V1 went from 8.732235404	to 5.69119973, and the importance of the duplicate1 variable is 4.28331581. So effectively the variable importance is split between the two correlated variables. 

What happens when you add another predictor that is also highly correlated with V1?

```{r}
simulated$duplicate2 <- simulated$V1 + rnorm(200) * .2
cor(simulated$duplicate2, simulated$V1)
```

```{r}
model3 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp3 <- varImp(model3, scale = FALSE) |> arrange(desc(Overall))
rfImp3
```

Adding a second correlated variable decreased the variable importance of V1 further.

(c) Use the cforest function in the party package to fit a random forest model using conditional inference trees. The party package function varimp can calculate predictor importance. The conditional argument of that function toggles between the traditional importance measure and the modified version described in Strobl et al. (2007). Do these importances show the same pattern as the traditional random forest model?

```{r}
library(party)


model_forest <- cforest(y ~ ., data = simulated)

varimp(model_forest, conditional = TRUE)
varimp(model_forest, conditional = FALSE)

```

These importances do show the same pattern of the random forest model where variables V1-V5 (except V3) have larger coefficients than V6-V10.

(d) Repeat this process with different tree models, such as boosted trees and Cubist. Does the same pattern occur?

First try Bagged Tree.

```{r}
library(tidyverse)
gbmGrid <- expand.grid(.interaction.depth = seq(1, 7, by = 2),
                       .n.trees = seq(100, 1000, by = 50),
                       .shrinkage = c(0.01, 0.1),
                       .n.minobsinnode = 1)

sim_x <- simulated |> select(-y)
gbmTune <- train(sim_x, simulated$y, method = "gbm", tuneGrid = gbmGrid, verbose = FALSE)

summary(gbmTune)
```

Next try cubist.

```{r}
cubistTuned <- train(sim_x, simulated$y, method = "cubist", verbose = FALSE)
varImp(cubistTuned)
```


For the both the bagged and cubist models variables 1-5 show up as the most important. 

### 8.2

Use a simulation to show tree bias with different granularities.

Basic regression trees "suffer from selection bias: predictors with a higher number of distinct values are favored over more granular predictors" [pg. 182]. The book notes that Loh and Shih explain that if a noisy variable has more splits than an informative variable - it still may be chosen over the informative variable just because it has higher resolution. 

To illustrate this concept, we can built a dataset that has predictors with different granularities, choose a subset to be used in creating a model and then see if the model picks the actual predictive predictors or just the noisy ones.

```{r}
library(rpart)
p1 <- rep(1,100)
x <- c(1,2)
p2 <- rep(x,50)
p3 <- rnorm(100, mean = 1000, sd = 200)
y <- p1 * p2 + rnorm(100)

df <- data.frame(p1 = p1, p2= p2, p3 = p3, y = y)

rpartTree <- rpart(y ~ ., data = df)
varImp(rpartTree)
```

The granularity of p1, there is only 1 distinct value, then for p2 there is only 2 distinct values and for p3 there are many distinct values. The variable importance of p3 is the highest. I think this is because it is picking up on the random noise. 

### 8.3

In stochastic gradient boosting the bagging fraction and learning rate will govern the construction of the trees as they are guided by the gradient. Although the optimal values of these parameters should be obtained through the tuning process, it is helpful to understand how the magnitudes of these parameters affect magnitudes of variable importance. Figure 8.24 provides the variable importance plots for boosting using two extreme values for the bagging fraction (0.1 and 0.9) and the learning rate (0.1 and 0.9) for the solubility data. The left-hand plot has both parameters set to 0.1, and the right-hand plot has both set to 0.9:

(a) Why does the model on the right focus its importance on just the first few of predictors, whereas the model on the left spreads importance across more predictors?

The model on the right focuses its importance on just the first few predictors whereas the model on the left spreads the importance across more predictors because of the differing bagging fraction and learning rates. The left hand has both set to 0.1 and the right hand has both set to 0.9. The learning rate is the "fraction of the current predicted value that is added to the previous iterations predicted value". The smaller the value, the greater the computation time. For the bagging fraction is the fraction of the training data that is randomly selected. With a smaller learning rate and a smaller bagging fraction, the variable importance may be more evenly distributed because there will be more trees created which gives more opportunity for a variable to be deemed "important". 

(b) Which model do you think would be more predictive of other samples?

I think the left model (low bagging fraction and low learning rate) would be more predictive of other samples because the low bagging fraction leads to a greater diversity in the trees and a low learning rate leads to more trees - it seems that this combination would have greater predictive power.

(c) How would increasing interaction depth affect the slope of predictor importance for either model in Fig. 8.24?

According to the book, "When regression tree are used as the base learner, simple gradient boosting for regression has two tuning parameters: tree depth and number of iterations. Tree depth in this context is also known as interaction depth, since each subsequential split can be thought of as a higher-level interaction term with all of the other previous split predictors." pg 205. Based on this, increasing the interaction depth means increasing the tree depth and doing so has the possibility of leading to overfitting the data. I would think that a smaller tree depth would lead to a more rapidly decreasing to 0 variable importance and a greater tree depth would have a more gradual decrease to 0. 

### 8.7

Refer to Exercises 6.3 and 7.5 which describe a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several tree-based models:

```{r}
data("ChemicalManufacturingProcess")
cmp_i <- knnImputation(ChemicalManufacturingProcess,k=3)

cmp_i2 <- cmp_i |> select(-Yield)

trans <- preProcess(cmp_i2, method = c("center", "scale", "YeoJohnson", "nzv"))
cmp_i2 <- predict(trans, cmp_i2)

cmp_i2$Yield <- cmp_i$Yield

cmp_i <- cmp_i2

i <- createDataPartition(cmp_i$Yield,
                                    p = 0.8,
                                    list = FALSE,
                                    times = 1)

x <- cmp_i[,colnames(cmp_i)[colnames(cmp_i) != 'Yield']]
y <- cmp_i[,colnames(cmp_i)[colnames(cmp_i) == 'Yield']]

data.train.x <- x[i,]
data.train.y <- y[i]
data.test.x <- x[-i,]
data.test.y <- y[-i]
```

(a) Which tree-based regression model gives the optimal resampling and test
set performance?

(b) Which predictors are most important in the optimal tree-based regression model? Do either the biological or process variables dominate the list? How do the top 10 important predictors compare to the top 10 predictors from the optimal linear and nonlinear models?

(c) Plot the optimal single tree with the distribution of yield in the terminal nodes. Does this view of the data provide additional knowledge about the biological or process predictors and their relationship with yield?

### Recommender Problem